Measuring matter effects in Neutron star - Black hole binaries with Advanced LIGO

### 2016-04-07 Paper/paper.tex
first results in intro

The goal of this paper is to probe the distinguishability of BHBH and NSBH
mergers, focusing at the effects of the BH mass, BH spin, and the tidal 
deformability parameter $\Lambda_\NS\sim(R/M)^5_\NS$. We also investigate the
accuracy with which aLIGO can measure $\Lambda_\NS$ itself. For both of 
these studies, we will use the zero-detuning high-power noise curve for the
aLIGO detectors. \prayush{In addition, we also study the effect of tuning aLIGO
over a narrow frequency range aimed at NSBH mergers.}
We use the models published in Ref.~\cite{Lackey:2013axa,Pannarale:2015jka} to
simulate NSBH merger waveforms.

>>>>

In this paper we study the measurability of neutron star's tidal deformability
from GWs emitted by {\it disruptive} NSBH mergers using aLIGO. We also probe how tidal 
effects affect the estimation of other binary parameters for the same class
of systems. This study improves upon previous work in many ways. First, we
include tidal effects during inspiral and merger in a consistent way, by using
the waveform model of Ref.~\cite{Lackey:2013axa} (henceforth ``LEA+''). Second,
we include the effect
of black hole spin on tidal GW signals, in addition to the effect of BH mass 
and NS's deformability itself. Third, we perform a fully Bayesian 
analysis, instead of using a Fisher matrix approximation. Fourth, we explore 
how our measurement errors decrease as we gain information from multiple 
(realistic) events.

We first probe the effect of not including tidal effects in parameter 
estimation templates on the recovery of non-tidal binary parameters, such as 
component masses and spins. This is the case of current and planned aLIGO
efforts.
To do so, we use the LEA+ model to generate a set of realistic signals, and
use that as data from which we estimate binary mass and spin parameters with
a Bayesian Monte-Carlo algorithm, using non-tidal (BBH) waveforms as filters.
% 
As in rest of the paper, we restrict here our parameter space to span 
mass-ratios $q:=\mbh/\mns\in[2,5]$, black hole spin $\chibh\in[-0.5, +0.75]$, 
and dimensionless neutron star tidal deformability 
$\lambdans:= G\left(\frac{c^2}{G \mns}\right)^5\lambda \in[500, 2000]$.
The mass of NS for signals (and not templates) is fixed to $1.35M_\odot$ 
throughout~\cite{stellarcollapsemass}, and is taken to be 
non-spinning~\cite{Miller:2014aaa}.
Each signal is rescaled to a range of SNR values $\rho\in[10, 70]$.
% 
We find that non-inclusion of tidal effects leads to systematic biases in
the measurement of non-tidal parameters. Consider first the binary chirp 
mass $\mchirp:=M\eta^{3/5}$ where $M$ is the binary total mass, and 
$\eta:=\mbh\mns/M^2$ is the dimensionless mass-ratio. We find that for
realistic (moderate) SNRs, i.e. $\rho\lesssim 30$, not including tidal
effects introduces a barely measurable bias that remains below $50\%$
of the statistical uncertainty of the $\mchirp$ measurement itself.
For higher signal strengths, we find that the underlying statistical
uncertainty decreases enough for the systematic bias to become comparable
to it for $\rho=50-70$ and higher. This is expected because chirp mass is fairly
accurately determined by the inspiral, where tidal effects are weak.
Next, we consider the mass-ratio $\eta$. We find that only if the NS is quite
deformable (with $\lambdans=2000$), {\it and} has a companion BH whose mass
is $\mbh\lesssim 4.5M_\odot$ (i.e. in the astrophysical ``mass-gap''~\cite{
Bailyn:1997xt,Kalogera:1996ci,Kreidberg:2012,Littenberg:2015tpa}),
will the systematic bias in $\eta$ measurement become comparable to its
statistical uncertainties at SNRs $\rho\leq 30$. For SNRs $\rho\gtrsim30$,
however, not including tidal effects is very likely to {\it significantly} bias
the mass-ratio estimate for NSBH binaries. Finally, we consider the spin on the
black hole $\chibh$. We find that $\chibh$ measurements are affected in a similar
way to $\eta$. By not including tidal effects, we compromise our measurement of
$\chibh$ significantly only if signal $\rho\gtrsim 30$. 
%
Overall, we learn an important thing. Bayesian parameter estimation algorithms
aimed at identifying NSBH binaries with low latency in order to alert 
electromagnetic (EM) observers in time to establish a possible coincident 
observation of a GRB~\cite{2012A&A...541A.155A,Singer:2014qca,Singer:2015ema,
Pankow:2015cra} can use BBH templates to look for NSBH systems with aLIGO.
This is so because the primary requirement of the EM follow-up effort is rapid
classification of a binary as an NSBH system, which can be achieved just as 
easily with BBH templates, purely on the basis of the smaller component's mass.
Another conclusion we draw is that detection searches are unaffected by the
choice of ignoring tidal effects in matched-filtering templates.
